---
title: "Sampling points in polygons and creatign climatic clusters"
author: "Mathew Rees"
date: "20/05/2024"
output: 
  html_notebook:
    number_sections: true
    code_folding: "hide"
    df_print: "paged"
    toc: true
    toc_depth: 3
    toc_float: true
---

In this script, I want to quickly show how to compare climatic clusters for a given region of interest.

First, let's load some libraries.

# Libraries

```{r}
library(tidyverse)
library(tidyterra)
library(sf)
library(sp)
library(raster)
library(RStoolbox)
library(rnaturalearth)
library(geodata)
library(viridis)
library(ade4)
library(factoextra)
library(alphahull)
library(mclust)
```


# Import polygons

We want to import a shapefile containing a polygon of the region of interest (ROI).

You can use this tool to define the ROI: https://www.openstreetmap.org/export#map=6/2.460/120.256


```{r}
Borneo_island <- ne_countries(scale="medium", returnclass = "sf") %>% 
  filter(admin %in% c("Indonesia", "Malaysia", "Brunei"))

ROI = c(xmin= 108.699, xmax = 119.355, ymin = -4.127, ymax = 7.580)

```

```{r}
Borneo <- st_crop(Borneo_island, st_bbox(ROI))
Borneo_sfc <- Borneo %>% st_as_sfc()
plot(Borneo_sfc)
```

Next I want to crop out the little part of West Sulawesi.

```{r}
cropped_extent <- c(xmin= 118, xmax = 119.355 , ymin = -4.144604, ymax = -1) %>% st_bbox() %>% st_as_sfc()
st_crs(cropped_extent) <- st_crs(Borneo_sfc)
```

```{r}
Borneo_sf <- st_difference(Borneo_sfc, cropped_extent) %>% st_as_sf()
ggplot() + geom_sf(data=Borneo_sf) + theme_classic()
```

Great. Now we can simply merge the outline of all three countries

```{r}
Borneo_united <- st_union(Borneo_sf)
ggplot() + geom_sf(data=Borneo_united) + theme_classic()
```


# Sampling points

Let's make some random sampling points that cover the island.

```{r}
p1 = st_sample(Borneo_united, 2000)
plot(st_geometry(Borneo_united))
plot(p1, add = TRUE, col="red")
```

Here are 2000 random sample points from Borneo. You can create as many or as few as you like.

```{r}
sample_pts<-as(p1, "Spatial") %>% coordinates() %>% `colnames<-`(c("longitude", "latitude")) %>% as.data.frame()
head(sample_pts)
```


# Bioclim variables

We can download Bioclim variables for our ROI

```{r}
bio <- worldclim_global(var = "bio", # "tmin", "tmax", "tavg", "prec", "wind", "vapr", or "bio"
                        res = 5, # resolution: 10, 5, 2.5, or 0.5 (minutes of a degree)
                        path = "./wc")
```


```{r}
plot(bio[[1]])
```

First, let's crop our `bio` object to the ROI.

```{r}
bio_cropped <- crop(bio, Borneo_sf)
plot(bio_cropped[[1]])
```

# RasterPCA

We can use the RSToolbox to build a PCA of Climate space


```{r}
bio_PCA <- rasterPCA(bio_cropped, nSamples = NULL, spca = T, maskCheck = T)

plot(bio_PCA$map[[1:4]])
```

See the variance explained by the PCA axes

```{r}
summary(bio_PCA$model)
```
So the first Axis explains 46.7%, the second 22.8%, third 13.9% and fourth 9.7%. Together they make up 93% of variance.

We can try to use a very simple k-means classifier with a predefined number of categories. Let's say 9.

See here : https://bleutner.github.io/RStoolbox/reference/unsuperClass.html 

```{r}
## Run classification
set.seed(12345)
unC <- unsuperClass(bio_PCA$map[[1:3]], nSamples = 10000, nClasses = 9, nStarts = 25)
unC

## Plots
colors <- rainbow(9)
plot(unC$map, col = colors, legend = FALSE, axes = T, box = FALSE)
##legend(1,1, legend = paste0("C",1:5), fill = colors, title = "Classes", horiz = TRUE,  bty = "n")
```


```{r}
## Return the distance of each pixel to each class centroid
unC2 <- unsuperClass(bio_PCA$map, nSamples = 10000, nClasses = 9, output = "distances")
unC2

ggR(unC2$map, 1:3, geom_raster = TRUE)
```


# Extracting values

We can combine these PCA axes to then select our points of interest and see where they fall

```{r}
stackPCA <- brick(bio_PCA$map[[1:4]])
```

and extract the point values

```{r}
test <- extract(stackPCA, sample_pts)
test2 <- extract(bio_cropped, sample_pts)
```

```{r}
plot(test[,1], test[,2])
```

Check these values

```{r}
sum(is.na(test2))
```

Some cells have NA values.
let's remove these before re-running the PCA manually

```{r}
test2.na.rm <- na.omit(test2)
```


```{r}
biopca2 <- dudi.pca(test2.na.rm[,-1], center = T, scale = T, scannf = F, nf = 4)
```

```{r}
fviz_pca(biopca2)
get_eig(biopca2)
fviz_eig(biopca2)
```

or a densityplot with an alpha hull contour
for this we need to remove any duplicated values in our pca

```{r}
BIO_PCs <- biopca2$li[,1:2]
BIO_PCs_unique <- BIO_PCs[!duplicated(BIO_PCs),]
alpha_obj <- ashape(BIO_PCs_unique, alpha=2)
```


```{r}
ggplot(data=biopca2$li, aes(Axis1, Axis2)) + 
  stat_density2d(aes(fill = after_stat(level)), geom = "polygon", colour="white", alpha =0.3) + 
  geom_point(size=0.2,alpha=0.2) +
  geom_segment(data=data.frame(alpha_obj$edges), aes(x = x1, y = y1, xend = x2, yend = y2), color="red") +
  theme_classic() + 
  scale_fill_viridis() + 
  labs(title = "Climatic envelope of Borneo",
              subtitle = "Red outline is alphahull and coloured shapes represent 2D density kernel") +
  theme(legend.position = "none")
```


# Classification

We can use the MClust package to make some density based clustering.

```{r}
library(mclust)

#Here I am using the first 3 axes only, but you can change this and see how it modifies the results
clusters <- Mclust(biopca2$li[,1:3])
```

```{r}
summary(clusters)
plot(clusters, "BIC")
```

Seems like 9 clusters is the best option

Let's plot these back onto the map

```{r}
ggplot() +
  geom_sf(data=Borneo_united) +
  geom_point(data = sample_pts[rownames(test2.na.rm),], aes(x=longitude, y=latitude, col=as.factor(clusters$classification))) + 
  theme_classic()
```

Compare with the K-means we did earlier

```{r}
plot(unC$map, col = colors, legend = FALSE, axes = T, box = FALSE)
```

Print session info

```{r}
sessionInfo()
```


